\section{Built-in Modules}

The modules described in this section are built into the interpreter.
They must be imported using
{\tt import}.
Some modules are not always available; it is a configuration option to
provide them.
Details are listed with the descriptions, but the best way to see if
a module exists in a particular implementation is to attempt to import
it.

\subsection{Built-in Module {\tt sys}}

This module provides access to some variables used or maintained by the
interpreter and to functions that interact strongly with the interpreter.
It is always available.
\begin{description}
\funcitem{argv}
The list of command line arguments passed to a {\Python} script.
{\tt sys.argv[0]}
is the script name.
If no script name was passed to the {\Python} interpreter,
{\tt sys.argv}
is empty.
\funcitem{exit}{n}
Exits from {\Python} with numeric exit status
{\tt n}.
This closes all open files and performs other cleanup-actions required by
the interpreter (but
{\em finally clauses}
of
{\tt try}
statements are not executed!).
\funcitem{modules}
Gives the list of modules that have already been loaded.
This can be manipulated to force reloading of modules and other tricks.
\funcitem{path}
A list of strings that specifies the search path for modules.
Initialized from the environment variable {\tt PYTHONPATH}, or an
installation-dependent default.
\funcitem{ps1,~ps2}
Strings specifying the primary and secondary prompt of the interpreter.
These are only defined if the interpreter is in interactive mode.
Their initial values in this case are
{\tt '>>> '}
and
{\tt '... '}.
\funcitem{stdin, stdout, stderr}
%.br
File objects corresponding to the interpreter's standard input, output
and error streams.
{\tt sys.stdin}
is used for all interpreter input except for scripts but including calls
to
{\tt input()}
and
{\tt raw\_input()}.
{\tt sys.stdout}
is used for the output of
{\tt print} and expression statements
and for the prompts of
{\tt input()}
and
{\tt raw\_input()}.
The interpreter's own prompts and its error messages are written to
stderr.
Assigning to
{\tt sys.stderr}
has no effect on the interpreter; it can be used to write error messages
to stderr using
{\tt print}.
\end{description}

\subsection{Built-in Module {\tt math}}

This module is always available.
It provides access to the mathematical functions defined by the C
standard.
They are:
{\tt acos(x)},
{\tt asin(x)},
{\tt atan(x)},
{\tt atan2(x,y)},
{\tt ceil(x)},
{\tt cos(x)},
{\tt cosh(x)},
{\tt exp(x)},
{\tt fabs(x)},
{\tt floor(x)},
%{\tt fmod(...)}  XXX not yet
%{\tt frexp(...)} XXX not yet
%{\tt ldexp(...)} XXX not yet
{\tt log(x)},
{\tt log10(x)},
%{\tt modf(...)}  XXX not yet
{\tt pow(x,y)},
{\tt sin(x)},
{\tt sinh(x)},
{\tt sqrt(x)},
{\tt tan(x)},
{\tt tanh(x)}.

It also defines two mathematical constants:
{\tt pi}
and
{\tt e}.

\subsection{Built-in Module {\tt time}}

This module provides various time-related functions.
It is always available.
Functions are:
\begin{description}
\funcitem{sleep}{secs}
Suspends execution for the given number of seconds.
\funcitem{time}{}
Returns the time in seconds since the Epoch (Thursday January 1,
00:00:00, 1970 UCT on \UNIX\ machines).
\end{description}

\noindent
In some versions (Amoeba, Mac) the following functions also exist:
\begin{description}
\funcitem{millisleep}{msecs}
Suspends execution for the given number of milliseconds.
\funcitem{millitimer}{}
Returns the number of milliseconds of real time elapsed since some point
in the past that is fixed per execution of the python interpreter (but
may change in each following run).
\end{description}

\noindent
The granularity of the milliseconds functions may be more than a
millisecond (100 msecs on Amoeba, 1/60 sec on the Mac).

\subsection{Built-in Module {\tt regexp}}

This module provides a regular expression matching operation.
It is always available.

The module defines a function and an exception:

\begin{description}

\funcitem{compile}{pattern}

Compile a regular expression given as a string into a regular
expression object.
The string must be an egrep-style regular expression;
this means that the characters
{\tt '(' ')' '*' '+' '?'\ '|' }\verb='^' '$'=
are special.
(It is implemented using Henry Spencer's regular expression matching
functions.)

excitem{error}{regexp.error}

Exception raised when a string passed to {\tt compile()} is not a
valid regular expression (e.g., unmatched parentheses) or when some other
error occurs during compilation or matching
(``no match found'' is not an error).

\end{description}

Compiled regular expression objects support a single method:

\begin{description}

\funcitem{exec}{str}

Find the first occurrence of the compiled regular expression in the
string {\tt str}.
The return value is a tuple of pairs specifying where a match was
found and where matches were found for subpatterns specified with
{\tt '('} and {\tt ')'} in the pattern.
If no match is found, an empty tuple is returned; otherwise the first
item of the tuple is a pair of slice indices into the search string
giving the match found.
If there were any subpatterns in the pattern, the returned tuple has an
additional item for each subpattern, giving the slice indices into the
search string where that subpattern was found.

\end{description}

For example:
\bcode\begin{verbatim}
>>> import regexp
>>> r = regexp.compile('--(.*)--')
>>> s = 'a--b--c'
>>> r.exec(s)
((1, 6), (3, 4))
>>> s[1:6] # The entire match
'--b--'
>>> s[3:4] # The subpattern
'b'
>>>
\end{verbatim}\ecode

\subsection{Built-in Module {\tt posix}}

This module provides access to operating system functionality that is
standardized by the C Standard and the POSIX standard (a thinly diguised
{\UNIX} interface).
It is available in all {\Python} versions except on the Macintosh.
Errors are reported exceptions.
It defines the following items:
\begin{description}
\funcitem{chdir}{path}
Changes the current directory to
{\tt path}.
\funcitem{chmod}{path, mode}
Change the mode of
{\tt path}
to the numeric
{\tt mode}.
\funcitem{environ}
A dictionary representing the string environment at the time
the interpreter was started.
(Modifying this dictionary does not affect the string environment of the
interpreter.)
For example,
{\tt posix.environ['HOME']}
is the pathname of your home directory, equivalent to
{\tt getenv("HOME")}
in C.
\excitem{error}{posix.error}
%.br
The exception raised when an POSIX function returns an error.
The value accompanying this exception is a pair containing the numeric
error code from
{\tt errno}
and the corresponding string, as would be printed by the C function
{\tt perror()}.
\funcitem{getcwd}{}
Returns a string representing the current working directory.
\funcitem{link}{src, dst}
Creates a hard link pointing to
{\tt src}
named
{\tt dst}.
\funcitem{listdir}{path}
Returns a list containing the names of the entries in the
directory.
The list is in arbitrary order.
It includes the special entries
{\tt '.'}
and
{\tt '..'}
if they are present in the directory.
\funcitem{mkdir}{path, mode}
Creates a directory named
{\tt path}
with numeric mode
{\tt mode}.
\funcitem{rename}{src, dst}
Renames the file or directory
{\tt src}
to
{\tt dst}.
\funcitem{rmdir}{path}
Removes the directory
{\tt path}.
\funcitem{stat}{path}
Performs a
{\em stat}
system call on the given path.
The return value is a tuple of at least 10 integers giving the most
important (and portable) members of the
{\em stat}
structure, in the order
{\tt st\_mode},
{\tt st\_ino},
{\tt st\_dev},
{\tt st\_nlink},
{\tt st\_uid},
{\tt st\_gid},
{\tt st\_size},
{\tt st\_atime},
{\tt st\_mtime},
{\tt st\_ctime}.
More items may be added at the end by some implementations.
\funcitem{system}{command}
Executes the command (a string) in a subshell.
This is implemented by calling the Standard C function
{\tt system()},
and has the same limitations.
Changes to
{\tt posix.environ},
{\tt sys.stdin}
etc. are not reflected in the environment of the executed command.
The return value is the exit status of the process as returned by
Standard C
{\tt system()}.
\funcitem{umask}{mask}
Sets the current numeric umask and returns the previous umask.
\funcitem{unlink}{path}
Unlinks the file
{\tt path}.
\funcitem{utimes(path, }{atime, mtime)}
%.br
Sets the access and modified time of the file to the given values.
(The second argument is a tuple of two items.)
\end{description}

The following functions are only available on systems that support
symbolic links:
\begin{description}
\funcitem{lstat}{path}
Like
{\tt stat()},
but does not follow symbolic links.
\funcitem{readlink}{path}
Returns a string representing the path to which the symbolic link
points.
\funcitem{symlink}{src, dst}
Creates a symbolic link pointing to
{\tt src}
named
{\tt dst}.
\end{description}

\subsection{Built-in Module {\tt stdwin}}

This module defines several new object types and functions that
provide access to the functionality of the Standard Window System
Interface, STDWIN [CWI report CR-R8817].
It is available on systems to which STDWIN has been ported (which is
most systems).
It is only available if the {\tt DISPLAY} environment variable is set
or an explicit `{\tt -display \it displayname}' argument is passed to
the interpreter.

Functions have names that usually resemble their C STDWIN counterparts
with the initial `w' dropped.
Points are represented by pairs of integers; rectangles
by pairs of points.
For a complete description of STDWIN please refer to the documentation
of STDWIN for C programmers (aforementioned CWI report).
\subsubsection{Functions Defined in Module {\tt stdwin}}

The following functions are defined in the {\tt stdwin} module:
\begin{description}
\funcitem{open}{title}
%.br
Opens a new window whose initial title is given by the string argument.
Returns a window object; window object methods are described below.%
\footnote{
The {\Python} version of STDWIN does not support draw procedures; all
drawing requests are reported as draw events.
}
\funcitem{getevent}{}
%.br
Waits for and returns the next event.
An event is returned as a triple: the first element is the event
type, a small integer; the second element is the window object to which
the event applies, or
{\tt None}
if it applies to no window in particular;
the third element is type-dependent.
Names for event types and command codes are defined in the standard
module
{\tt stdwinevent}.
\funcitem{setdefwinpos}{h, v}
%.br
Sets the default window position.
\funcitem{setdefwinsize}{width, height}
%.br
Sets the default window size.
\funcitem{menucreate}{title}
%.br
Creates a menu object referring to a global menu (a menu that appears in
all windows).
Methods of menu objects are described below.
\funcitem{fleep}{}
%.br
Causes a beep or bell (or perhaps a `visual bell' or flash, hence the
name).
\funcitem{message}{string}
%.br
Displays a dialog box containing the string.
The user must click OK before the function returns.
\funcitem{askync}{prompt, default}
%.br
Displays a dialog that prompts the user to answer a question with yes or
no.
The function returns 0 for no, 1 for yes.
If the user hits the Return key, the default (which must be 0 or 1) is
returned.
If the user cancels the dialog, the
{\tt KeyboardInterrupt}
exception is raised.
\funcitem{askstr}{prompt, default}
%.br
Displays a dialog that prompts the user for a string.
If the user hits the Return key, the default string is returned.
If the user cancels the dialog, the
{\tt KeyboardInterrupt}
exception is raised.
\funcitem{askfile}{prompt, default, new}
%.br
Asks the user to specify a filename.
If
{\tt new}
is zero it must be an existing file; otherwise, it must be a new file.
If the user cancels the dialog, the
{\tt KeyboardInterrupt}
exception is raised.
\funcitem{setcutbuffer}{i, string}
%.br
Stores the string in the system's cut buffer number
{\tt i},
where it can be found (for pasting) by other applications.
On X11, there are 8 cut buffers (numbered 0..7).
Cut buffer number 0 is the `clipboard' on the Macintosh.
\funcitem{getcutbuffer}{i}
%.br
Returns the contents of the system's cut buffer number
{\tt i}.
\funcitem{rotatebutbuffers}{n}
%.br
On X11, this rotates the 8 cut buffers by
{\tt n}.
Ignored on the Macintosh.
\funcitem{getselection}{i}
%.br
Returns X11 selection number
{\tt i.}
Selections are not cut buffers.
Selection numbers are defined in module
{\tt stdwinevents}.
Selection {\tt WS\_PRIMARY} is the
{\em primary}
selection (used by
xterm,
for instance);
selection {\tt WS\_SECONDARY} is the
{\em secondary}
selection; selection {\tt WS\_CLIPBOARD} is the
{\em clipboard}
selection (used by
xclipboard).
On the Macintosh, this always returns an empty string.
\funcitem{resetselection}{i}
%.br
Resets selection number
{\tt i},
if this process owns it.
(See window method
{\tt setselection()}).
\funcitem{baseline}{}
%.br
Return the baseline of the current font (defined by STDWIN as the
vertical distance between the baseline and the top of the
characters).%
\footnote{
There is no way yet to set the current font.
This will change in a future version.
}
\funcitem{lineheight}{}
%.br
Return the total line height of the current font.
\funcitem{textbreak}{str, width}
%.br
Return the number of characters of the string that fit into a space of
{\tt width}
bits wide when drawn in the curent font.
\funcitem{textwidth}{str}
%.br
Return the width in bits of the string when drawn in the current font.
\end{description}

\subsubsection{Window Object Methods}

Window objects are created by
{\tt stdwin.open()}.
There is no explicit function to close a window; windows are closed when
they are garbage-collected.
Window objects have the following methods:
\begin{description}
\funcitem{begindrawing}{}
Returns a drawing object, whose methods (described below) allow drawing
in the window.
\funcitem{change}{rect}
Invalidates the given rectangle; this may cause a draw event.
\funcitem{gettitle}{}
Returns the window's title string.
\funcitem{getdocsize}{}
\begin{sloppypar}
Returns a pair of integers giving the size of the document as set by
{\tt setdocsize()}.
\end{sloppypar}
\funcitem{getorigin}{}
Returns a pair of integers giving the origin of the window with respect
to the document.
\funcitem{getwinsize}{}
Returns a pair of integers giving the size of the window.
\funcitem{menucreate}{title}
Creates a menu object referring to a local menu (a menu that appears
only in this window).
Methods menu objects are described below.
\funcitem{scroll}{rect,~point}
Scrolls the given rectangle by the vector given by the point.
\funcitem{setwincursor}{name}
\begin{sloppypar}
Sets the window cursor to a cursor of the given name.
It raises the
{\tt Runtime\-Error}
exception if no cursor of the given name exists.
Suitable names are
{\tt 'ibeam'},
{\tt 'arrow'},
{\tt 'cross'},
{\tt 'watch'}
and
{\tt 'plus'}.
On X11, there are many more (see
{\tt <X11/cursorfont.h>}).
\end{sloppypar}
\funcitem{setdocsize}{point}
Sets the size of the drawing document.
\funcitem{setorigin}{point}
Moves the origin of the window to the given point in the document.
\funcitem{setselection}{i, str}
Attempts to set X11 selection number
{\tt i}
to the string
{\tt str}.
(See stdwin method
{\tt getselection()}
for the meaning of
{\tt i}.)
Returns true if it succeeds.
If it succeeds, the window ``owns'' the selection until
(a) another applications takes ownership of the selection; or
(b) the window is deleted; or
(c) the application clears ownership by calling
{\tt stdwin.resetselection(i)}.
When another application takes ownership of the selection, a
{\tt WE\_LOST\_SEL}
event is received for no particular window and with the selection number
as detail.
Ignored on the Macintosh.
\funcitem{settitle}{title}
Sets the window's title string.
\funcitem{settimer}{dsecs}
Schedules a timer event for the window in
{\tt dsecs/10}
seconds.
\funcitem{show}{rect}
Tries to ensure that the given rectangle of the document is visible in
the window.
\funcitem{textcreate}{rect}
Creates a text-edit object in the document at the given rectangle.
Methods of text-edit objects are described below.
\end{description}

\subsubsection{Drawing Object Methods}

Drawing objects are created exclusively by the window method
{\tt begindrawing()}.
Only one drawing object can exist at any given time; the drawing object
must be deleted to finish drawing.
No drawing object may exist when
{\tt stdwin.getevent()}
is called.
Drawing objects have the following methods:
\begin{description}
\funcitem{box}{rect}
Draws a box around a rectangle.
\funcitem{circle}{center, radius}
%.br
Draws a circle with given center point and radius.
\funcitem{elarc(center, (rh, rv), }{a1, a2)}
%.br
Draws an elliptical arc with given center point.
{\tt (rh, rv)}
gives the half sizes of the horizontal and vertical radii.
{\tt (a1, a2)}
gives the angles (in degrees) of the begin and end points.
0 degrees is at 3 o'clock, 90 degrees is at 12 o'clock.
\funcitem{erase}{rect}
Erases a rectangle.
\funcitem{invert}{rect}
Inverts a rectangle.
\funcitem{line}{p1, p2}
Draws a line from point
{\tt p1}
to
{\tt p2}.
\funcitem{paint}{rect}
Fills a rectangle.
\funcitem{text}{p, str}
Draws a string starting at point p (the point specifies the
top left coordinate of the string).
\funcitem{shade}{rect, percent}
%.br
Fills a rectangle with a shading pattern that is about
{\tt percent}
percent filled.
\funcitem{xorline}{p1, p2}
Draws a line in XOR mode.
\funcitem{baseline(), lineheight(), textbreak(), textwidth}{}
%.br
These functions are similar to the corresponding functions described
above for the
{\tt stdwin}
module, but use the current font of the window instead of the (global)
default font.
\end{description}

\subsubsection{Menu Object Methods}

A menu object represents a menu.
The menu is destroyed when the menu object is deleted.
The following methods are defined:
\begin{description}
\funcitem{additem}{text, shortcut}
%.br
Adds a menu item with given text.
The shortcut must be a string of length 1, or omitted (to specify no
shortcut).
\funcitem{setitem}{i, text}
Sets the text of item number
{\tt i}.
\funcitem{enable}{i, flag}
Enables or disables item
{\tt i}.
\funcitem{check}{i, flag}
Sets or clears the
{\em check mark}
for item
{\tt i}.
\end{description}

\subsubsection{Text-edit Object Methods}

A text-edit object represents a text-edit block.
For semantics, see the STDWIN documentation for C programmers.
The following methods exist:
\begin{description}
\funcitem{arrow}{code}
Passes an arrow event to the text-edit block.
The
{\tt code}
must be one of
{\tt WC\_LEFT},
{\tt WC\_RIGHT},
{\tt WC\_UP}
or
{\tt WC\_DOWN}
(see module
{\tt stdwinevents}).
\funcitem{draw}{rect}
Passes a draw event to the text-edit block.
The rectangle specifies the redraw area.
\funcitem{event}{type, window, detail}
%.br
Passes an event gotten from
{\tt stdwin.getevent()}
to the text-edit block.
Returns true if the event was handled.
\funcitem{getfocus}{}
Returns 2 integers representing the start and end positions of the
focus, usable as slice indices on the string returned by
{\tt getfocustext()}.
\funcitem{getfocustext}{}
Returns the text in the focus.
\funcitem{getrect}{}
Returns a rectangle giving the actual position of the text-edit block.
(The bottom coordinate may differ from the initial position because
the block automatically shrinks or grows to fit.)
\funcitem{gettext}{}
Returns the entire text buffer.
\funcitem{move}{rect}
Specifies a new position for the text-edit block in the document.
\funcitem{replace}{str}
Replaces the focus by the given string.
The new focus is an insert point at the end of the string.
\funcitem{setfocus}{i,~j}
Specifies the new focus.
Out-of-bounds values are silently clipped.
\end{description}

\subsubsection{Example}

Here is a simple example of using STDWIN in Python.
It creates a window and draws the string ``Hello world'' in the top
left corner of the window.
The window will be correctly redrawn when covered and re-exposed.
The program quits when the close icon or menu item is requested.
\bcode\begin{verbatim}
import stdwin
from stdwinevents import *

def main():
    mywin = stdwin.open('Hello')
    #
    while 1:
        (type, win, detail) = stdwin.getevent()
        if type = WE_DRAW:
            draw = win.begindrawing()
            draw.text((0, 0), 'Hello, world')
            del draw
        elif type = WE_CLOSE:
            break

main()
\end{verbatim}\ecode

\subsection{Built-in Module {\tt amoeba}}

This module provides some object types and operations useful for
Amoeba applications.
It is only available on systems that support Amoeba operations.
RPC errors and other Amoeba errors are reported as the exception
{\tt amoeba.error = 'amoeba.error'}.
The module
{\tt amoeba}
defines the following items:
\begin{description}
\funcitem{name\_append}{path,~cap}
%.br
Stores a capability in the Amoeba directory tree.
Arguments are the pathname (a string) and the capability (a capability
object as returned by
{\tt name\_lookup()}).
\funcitem{name\_delete}{path}
%.br
Deletes a capability from the Amoeba directory tree.
Argument is the pathname.
\funcitem{name\_lookup}{path}
%.br
Looks up a capability.
Argument is the pathname.
Returns a
{\em capability}
object, to which various interesting operations apply, described below.
\funcitem{name\_replace}{path,~cap}
%.br
Replaces a capability in the Amoeba directory tree.
Arguments are the pathname and the new capability.
(This differs from
{\tt name\_append()}
in the behavior when the pathname already exists:
{\tt name\_append()}
finds this an error while
{\tt name\_replace()}
allows it, as its name suggests.)
\funcitem{capv}
A table representing the capability environment at the time the
interpreter was started.
(Alas, modifying this table does not affect the capability environment
of the interpreter.)
For example,
{\tt amoeba.capv['ROOT']}
is the capability of your root directory, similar to
{\tt getcap("ROOT")}
in C.
\excitem{error}{amoeba.error}
%.br
The exception raised when an Amoeba function returns an error.
The value accompanying this exception is a pair containing the numeric
error code and the corresponding string, as returned by the C function
{\tt err\_why()}.
\funcitem{timeout}{msecs}
%.br
Sets the transaction timeout, in milliseconds.
Returns the previous timeout.
Initially, the timeout is set to 2 seconds by the {\Python} interpreter.
\end{description}

\subsubsection{Capability Operations}

Capabilities are written in a convenient ASCII format, also used by the
Amoeba utilities
{\em c2a}(U)
and
{\em a2c}(U).
For example:
\bcode\begin{verbatim}
>>> amoeba.name_lookup('/profile/cap')
aa:1c:95:52:6a:fa/14(ff)/8e:ba:5b:8:11:1a
>>>
\end{verbatim}\ecode
The following methods are defined for capability objects.
\begin{description}
\funcitem{dir\_list}{}
Returns a list of the names of the entries in an Amoeba directory.
\funcitem{b\_read}{offset, maxsize}
%.br
Reads (at most)
{\tt maxsize}
bytes from a bullet file at offset
{\tt offset.}
The data is returned as a string.
EOF is reported as an empty string.
\funcitem{b\_size}{}
Returns the size of a bullet file.
\funcitem{dir\_append(), dir\_delete(), dir\_lookup(), dir\_replace}{}
%.br
\itembreak
Like the corresponding
{\tt name\_*}
functions, but with a path relative to the capability.
(For paths beginning with a slash the capability is ignored, since this
is the defined semantics for Amoeba.)
\funcitem{std\_info}{}
Returns the standard info string of the object.
\funcitem{tod\_gettime}{}
Returns the time (in seconds since the Epoch, in UCT, as for POSIX) from
a time server.
\funcitem{tod\_settime}{t}
Sets the time kept by a time server.
\end{description}

\subsection{Built-in Module {\tt audio}}

This module provides rudimentary access to the audio I/O device
{\tt /dev/audio}
on the Silicon Graphics Personal IRIS; see audio(7).
It supports the following operations:
\begin{description}
\funcitem{setoutgain}{n}
Sets the output gain (0-255).
\funcitem{getoutgain}{}
Returns the output gain.
\funcitem{setrate}{n}
Sets the sampling rate: 1=32K/sec, 2=16K/sec, 3=8K/sec.
\funcitem{setduration}{n}
Sets the `sound duration' in units of 1/100 seconds.
\funcitem{read}{n}
Reads a chunk of
{\tt n}
sampled bytes from the audio input (line in or microphone).
The chunk is returned as a string of length n.
Each byte encodes one sample as a signed 8-bit quantity using linear
encoding.
This string can be converted to numbers using {\tt chr2num()} described
below.
\funcitem{write}{buf}
Writes a chunk of samples to the audio output (speaker).
\end{description}

These operations support asynchronous audio I/O:
\begin{description}
\funcitem{start\_recording}{n}
%.br
Starts a second thread (a process with shared memory) that begins reading
{\tt n}
bytes from the audio device.
The main thread immediately continues.
\funcitem{wait\_recording}{}
%.br
Waits for the second thread to finish and returns the data read.
\funcitem{stop\_recording}{}
%.br
Makes the second thread stop reading as soon as possible.
Returns the data read so far.
\funcitem{poll\_recording}{}
%.br
Returns true if the second thread has finished reading (so
{\tt wait\_recording()} would return the data without delay).
\item[{\tt start\_playing(chunk)}, {\tt wait\_playing()},
{\tt stop\_playing()}, {\tt poll\_playing()}]
%.br
\begin{sloppypar}
Similar but for output.
{\tt stop\_playing()}
returns a lower bound for the number of bytes actually played (not very
accurate).
\end{sloppypar}
\end{description}

The following operations do not affect the audio device but are
implemented in C for efficiency:
\begin{description}
\funcitem{amplify}{buf, f1, f2}
%.br
Amplifies a chunk of samples by a variable factor changing from
{\tt f1}/256 to {\tt f2}/256.
Negative factors are allowed.
Resulting values that are to large to fit in a byte are clipped.
\funcitem{reverse}{buf}
%.br
Returns a chunk of samples backwards.
\funcitem{add}{buf1, buf2}
%.br
Bytewise adds two chunks of samples.
Bytes that exceed the range are clipped.
If one buffer shorter, it is assumed to be padded with zeros.
\funcitem{chr2num}{buf}
%.br
Converts a string of sampled bytes as returned by {\tt read()} into
a list containing the numeric values of the samples.
\funcitem{num2chr}{list}
%.br
\begin{sloppypar}
Converts a list as returned by
{\tt chr2num()}
back to a buffer acceptable by
{\tt write()}.
\end{sloppypar}
\end{description}

\subsection{Built-in Module {\tt gl}}

This module provides access to the Silicon Graphics
{\em Graphics Library}.
It is available only on Silicon Graphics machines.

{\bf Warning:}
Some illegal calls to the GL library cause the {\Python} interpreter to dump
core.
In particular, the use of most GL calls is unsafe before the first
window is opened.

The module is too large to document here in its entirety, but the
following should help you to get started.
The parameter conventions for the C functions are translated to {\Python} as
follows:

\begin{itemize}
\item
All (short, long, unsigned) int values are represented by {\Python}
integers.
\item
All float and double values are represented by {\Python} floating point
numbers.
In most cases, {\Python} integers are also allowed.
\item
All arrays are represented by one-dimensional {\Python} lists.
In most cases, tuples are also allowed.
\item
\begin{sloppypar}
All string and character arguments are represented by {\Python} strings,
for instance,
{\tt winopen('Hi~There!')}
and
{\tt rotate(900,~'z')}.
\end{sloppypar}
\item
All (short, long, unsigned) integer arguments or return values that are
only used to specify the length of an array argument are omitted.
For example, the C call
\bcode\begin{verbatim}
lmdef(deftype, index, np, props)
\end{verbatim}\ecode
is translated to {\Python} as
\bcode\begin{verbatim}
lmdef(deftype, index, props)
\end{verbatim}\ecode
\item
Output arguments are omitted from the argument list; they are
transmitted as function return values instead.
If more than one value must be returned, the return value is a tuple.
If the C function has both a regular return value (that is not omitted
because of the previous rule) and an output argument, the return value
comes first in the tuple.
Examples: the C call
\bcode\begin{verbatim}
getmcolor(i, &red, &green, &blue)
\end{verbatim}\ecode
is translated to {\Python} as
\bcode\begin{verbatim}
red, green, blue = getmcolor(i)
\end{verbatim}\ecode
\end{itemize}

The following functions are non-standard or have special argument
conventions:
\begin{description}
\funcitem{varray}{}
Equivalent to but faster than a number of
{\tt v3d()}
calls.
The argument is a list (or tuple) of points.
Each point must be a tuple of coordinates (x, y, z) or (x, y).
The points may be 2- or 3-dimensional but must all have the
same dimension.
Float and int values may be mixed however.
The points are always converted to 3D double precision points
by assuming z=0.0 if necessary (as indicated in the man page),
and for each point
{\tt v3d()}
is called.
\funcitem{nvarray}{}
Equivalent to but faster than a number of
{\tt n3f}
and
{\tt v3f}
calls.
The argument is an array (list or tuple) of pairs of normals and points.
Each pair is a tuple of a point and a normal for that point.
Each point or normal must be a tuple of coordinates (x, y, z).
Three coordinates must be given.
Float and int values may be mixed.
For each pair,
{\tt n3f()}
is called for the normal, and then
{\tt v3f()}
is called for the point.
\funcitem{vnarray}{}
Similar to
{\tt nvarray()}
but the pairs have the point first and the normal second.
\funcitem{nurbssurface}{s\_k[], t\_k[], ctl[][], s\_ord, t\_ord, type}
%.br
\itembreak
Defines a nurbs surface.
The dimensions of
{\tt ctl[][]}
are computed as follows:
{\tt [len(s\_k)~-~s\_ord]},
{\tt [len(t\_k)~-~t\_ord]}.
\funcitem{nurbscurve}{knots, ctlpoints, order, type}
%.br
Defines a nurbs curve.
The length of ctlpoints is
{\tt len(knots)~-~order}.
\funcitem{pwlcurve}{points, type}
%.br
Defines a piecewise-linear curve.
{\tt points}
is a list of points.
{\tt type}
must be
{\tt N\_ST}.
\funcitem{pick(n), select}{n}
%.br
The only argument to these functions specifies the desired size of the
pick or select buffer.
\funcitem{endpick(), endselect}{}
%.br
These functions have no arguments.
They return a list of integers representing the used part of the
pick/select buffer.
No method is provided to detect buffer overrun.
\end{description}

Here is a tiny but complete example GL program in {\Python}:
\bcode\begin{verbatim}
import gl, GL, time

def main():
    gl.foreground()
    gl.prefposition(500, 900, 500, 900)
    w = gl.winopen('CrissCross')
    gl.ortho2(0.0, 400.0, 0.0, 400.0)
    gl.color(GL.WHITE)
    gl.clear()
    gl.color(GL.RED)
    gl.bgnline()
    gl.v2f(0.0, 0.0)
    gl.v2f(400.0, 400.0)
    gl.endline()
    gl.bgnline()
    gl.v2f(400.0, 0.0)
    gl.v2f(0.0, 400.0)
    gl.endline()
    time.sleep(5)

main()
\end{verbatim}\ecode

\subsection{Built-in Module {\tt pnl}}

This module provides access to the
{\em Panel Library}
built by NASA Ames (to get it, send e-mail to
{\tt panel-requ...@nas.nasa.gov}).
All access to it should be done through the standard module
{\tt panel},
which transparantly exports most functions from
{\tt pnl}
but redefines
{\tt pnl.dopanel()}.

{\bf Warning:}
the {\Python} interpreter will dump core if you don't create a GL window
before calling
{\tt pnl.mkpanel()}.

The module is too large to document here in its entirety.
